---
title: "GHP 538 Final Proj"
author: "Angela Stegmuller"
date: "4/26/2023"
output:
  word_document: default
  html_document: default
  pdf_document: default
---



```{r}
library(sf) # manipulate simple (spatial) features
library(spdep) # spatial statistics
library(dplyr) # data management
library(ggplot2) # visualization and mapping
library(ggsn) # separate package for ggplot() north symbol & scale bar
library(viridis) # color palette
library(patchwork) # for plotting multiple ggplots together
library(rgeoda) # for bivariate moran's I and other functions
library(data.table) #making df
library(lubridate) #for dates
library(tigris) #county shapefile

```


create path
```{r}
path_to_data <- "/Users/angelastegmuller/Library/CloudStorage/OneDrive-HarvardUniversity/Spring 2023/GHP 538/final_proj"
```

Download LA shapefile from tigris package
```{r}
# download Louisiana counties shapefile
la_counties_sp <- counties(state = "LA", cb = TRUE)
summary(la_counties_sp$GEOID)
la_counties_sp$fips<-as.numeric(la_counties_sp$GEOID)

# convert to sf object
la_counties_sf <- st_as_sf(la_counties_sp)


# plot Louisiana counties
ggplot() +
  geom_sf(data = la_counties_sf)
```
Check CRS
```{r}
sf::st_crs(la_counties_sf) 

```

Creating weights using Queen's contiguity. 
```{r}
# (1a) Create a neighbors list based on Queen's contiguity
queens_nb <- spdep::poly2nb(pl = la_counties_sf, 
                            snap = 0.1, # edges < snap distance are "neighbors" this makes sure that edges are touching. 0.1 should be good 
                            queen = TRUE) # queen=TRUE is default
```
Now trying with centroid-based
```{r}
# Create a new object with centroids for our polygons
coords <- sf::st_centroid(st_geometry(la_counties_sf)) # This provides *just* the geometry
                                                  
head(coords)
class(coords)

plot(la_counties_sf[0])
plot(coords, add=TRUE, col="blue")
```
creating weights 
```{r}
lw_queens <- spdep::nb2listw(neighbours = queens_nb, 
                             style = 'W', # row-standardized 
                             zero.policy = TRUE) # this is default: permit creation of
                                                 #   list w object in cases where some 
                                                 #   units have no neighbors. works for islands 

class(lw_queens)
names(lw_queens)
lw_queens

```

Plot preterm birth

```{r}
ptb<-read.csv("/Users/angelastegmuller/Library/CloudStorage/OneDrive-HarvardUniversity/Spring 2023/GHP 538/final_proj/ptb_LA_2018-2021_Average.csv", stringsAsFactors = FALSE) 

names(ptb)
ptb<-ptb[-1,]

ptb$county_fip<-as.numeric(ptb$fipscode)

min(ptb$Preterm.Birth.Rate)
max(ptb$Preterm.Birth.Rate)

# join the preterm birth rates data with the parish shapefile
parish_preterm <- left_join(la_counties_sp, ptb, by = c("fips" = "county_fip"))
centroids <- st_centroid(parish_preterm) 
class(centroids) # It's still an sf object with a dataframe
plot(centroids[0])

plot_ptb <-ggplot(parish_preterm) +
  geom_sf(aes(fill = Preterm.Birth.Rate)) +
  scale_fill_gradient(low = "white", high = "#A46A80") +
  labs(title = "Preterm Birth Rates by Parish",
       fill = "Preterm Birth Rate",
       caption = "Source: March of Dimes") 
plot_ptb
```
Now Global Moran's I
```{r}
### Global Moran's I statistic 
moran_stat <- spdep::moran.test(x = parish_preterm$Preterm.Birth.Rate, listw = lw_queens)
moran_stat

# save the estimate for later
moran_estimate <- as.numeric(moran_stat$estimate[1])

# Create a Moran scatterplot

# Standardize the preterm birth rate values as a Z-score before creating scatterplot
parish_preterm$Preterm.Birth.Rate_z <- (parish_preterm$Preterm.Birth.Rate - mean(parish_preterm$Preterm.Birth.Rate))/sd(parish_preterm$Preterm.Birth.Rate)
summary(parish_preterm$Preterm.Birth.Rate_z)


moran_scatter<-spdep::moran.plot(x = parish_preterm$Preterm.Birth.Rate_z, listw = lw_queens,   
                       xlim=c(-2.4, 3.5), # widen a bit so that labels fit
                       labels=as.character(parish_preterm$Parish), # add labels
                       xlab="Preterm birth (z)", 
                       ylab="Spatially lagged preterm birth (z)")
```
Trying Moran's I. Is preterm birth completely spatially random in state of Louisiana? 

H0: Preterm birth rate is completely spatially random in Louisiana

Reject then null hypothesis with Global Moran's I of 0.369 and p-value <0.001, there is evidence of strong positive autocorrelation, meaning that similar preterm birth rates are close ot each other. 

Parishes like Tensas, East Carroll, and Red River contribute to positive spatial autocorrelation and Cameron contributes to negative (dissimilar values nearby)

Now onto local Moran's I
```{r}
# Scale the GEM value as a z-score
parish_preterm$Preterm.Birth.Rate_v <- as.vector(scale(parish_preterm$Preterm.Birth.Rate))

# Local moran functions from spdep
local_moran <- spdep::localmoran(x = parish_preterm$Preterm.Birth.Rate_v, 
                                 listw = lw_queens, # use our queens contiguity listw
                                 alternative = "two.sided")

local_moran <- spdep::localmoran_perm(x = parish_preterm$Preterm.Birth.Rate_v, 
                                      nsim=999,
                                 listw = lw_queens, # use our queens contiguity listw
                                 alternative = "two.sided")
summary(local_moran)
class(local_moran)
```
```{r}
hist(spdep::p.adjustSP(p = local_moran[,5], nb = queens_nb, method = "fdr"), 
     breaks = 15, main="adjust FDR")
# Histogram of current results
hist(spdep::p.adjustSP(p = local_moran[,5], nb = queens_nb, method = "none"), 
     breaks = 15, main="p-val; no adjustment")

```